---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# dfprint

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of `dfprint` is to provide a nicer print method for data frames,
inspired by tibbles.

## Installation

You can install `dfprint` from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("jchrom/dfprint")
```
## How does it print

`dfprint` includes text formatting, additional information such as type/class
of variables, robust printing of list columns, better use of screen-estate,
and better handling of some exotic phenomena such as list columns with symbol elements or 2-d columns.

```{r, include = FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(width = 80)
```

```{r example}
library(dfprint)
```

The readme files on GitHub do not seem to support ANSI formatting, so you better
be viewing this page on the package website.

```{r}
set.seed(1)
iris2 = iris[sample.int(nrow(iris), nrow(iris)), ]
iris2
```

See `vignette("print-data-frames")` for more examples and additional info.

## Dependencies

Only what comes with base R. If your terminal does not fully support ANSI strings,
use `options()` to disable offending features, eg.
`options(dfprint_strong = FALSE)`.
