---
title: "Printing data frames"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(width = 80)
```

**"But data frames already have a print method!"**

They do. But it falls short in a few few well-known respects which, at least
in part, motivated the "Tidyverse" folks to reimagine data frames altogether.
I'm going to go through these issues in order (of how much they annoy me).

**1. Large data frames print poorly.** Tinkering with `[[` and `head()` is not
fun. Lists of data frames may even hang R when printed in full.

**2. Lack of metadata.** These are typically obtained by `str()`, `dim()` and
`object.size()`, at the cost of additional typing and scrolling, as the original
output gets displaced by the subsequent calls.

**3. Too plain.** Less is not always more. ANSI formatting would help to tell 
apart factor from character, lists from atomic vectors etc.

**4. Valid inputs can produce an error.** For example:

```{r, error = TRUE}
li = list(as.symbol("a"), as.symbol("b"))
df = data.frame(a = 1:2)

df$li = li

df
```

The above is a data frame that cannot be printed because its list column `li`
includes symbol elements. Not a frequent use case, but should be handled as well.

**5. And then there is this:** Data frames can have columns which themselves are
dataframes (or matrices or arrays). One place where you can see this is when you
parse JSON data structures, eg. using `jsonlinte::fromJSON()`. Can you tell from
the printout below which columns belong to the data frame column?

```{r}
df_parent = data.frame(x = 1:2, child.y = letters[1:2])
df_child  = data.frame(y = LETTERS[1:2], z = 2:3)

# Add child data frame as a column to the parent.
df_parent$child = df_child

# Check the number of columns
ncol(df_parent)

# Print out the result.
df_parent
```

There is no way of telling that the first `child.y` is a regular column, while
the rest belongs to the child data frame. The "dot syntax" just does not help.

Burn it all down
----------------

Using *[tibbles](https://tibble.tidyverse.org/articles/tibble.html#tibbles-vs-data-frames)*
solves these problems nicely, but not for data frames. Thus, if you want to use
their nice print method, you have to first coerce your data frame into a tibble.
This is something you may not always want, be it because tibbles behave
in a slightly different way than data frames, or because some R functions expect
a data frame. If you don't mind that, you can do this for usage in interactive
sessions:

```r
print.data.frame = function(df) print(tibble::as_tibble(df))
rm(print.data.frame) # when you no longer need it
```

Note that this will fail when POSIXlt columns are present, and will drop row
names.

If you don't want to go the tibble route, the *dfprint* package is your friend.
Just load it in the usual way, and data frames will print nicely until you
detach it.

```{r}
library(dfprint)
```

Ah yes, it does overload the base print method for data frames.

Is it worth it?
---------------

I let you decide. In the meantime, let's have a look at how it prints. Here is
the `iris` data set with scrambled rows. I scrambled the rows so you can see
the factor ordering properly:

```{r}
set.seed(1)
iris2 = iris[sample.int(nrow(iris)), ]
iris2
```

Things to point out:

* The number of rows and columns is printed at the top, along with the amount
  of RAM allocated.
  
* Column headers include object type. For objects with a "class" attribute set
  (such as factors or datetimes), class is shown instead of type. Datetime
  columns (POSIXct and POSIXlt) also show the timezone.
  
* Factors display their underlying integer values to make their ordering a bit
  more obvious. This may be useful for eg. plotting.
  
* The footer is telling you how many rows are displayed. If you need more, use
  the the `n` argument, eg, `print(iris2, n = 20)`. Use `Inf` to print all rows.
  
When a data frame is bit wider, the names and summaries of the overflowing
columns are displayed below.

```{r}
set.seed(1)
iris3 = iris2[, sample.int(ncol(iris2), 8, replace = TRUE)]
iris3
```

Row names are preserved:

```{r}
mtcars[1:4, ]
```

And what about those failed examples from before? Here is the "unprintable"
data frame with some additional list columns added:

```{r}
df$li2 = list(mtcars, letters)
df$li3 = list(mean, ~x)
df
```

Prints out no problem. List columns include summaries of their elements instead
of the values themselves, which saves space and is often more informative.

Now for the data frame column - and let's also add a matrix column:

```{r}
df_parent$m = matrix(1:4, ncol = 2)
df_parent
```

The `$` and `[` make it obvious that these columns belong to 2-d structures. Note
that the number of columns stated above is correct, because `child` (not
`child.y`) and `m` are each a single 2-d column.

Options
-------

If your terminal does not fully support ANSI formatting, you can turn selected
features off.

```{r}
options(dfprint_strong = FALSE)

# Column headers are no longer bold:
mtcars
```

Use `dfprint_italic = FALSE` and `dfprint_subtle = FALSE` to disable individual
features, or `dfprint_enable = FALSE` to revert back to base print method.

Using dfprint in other packages
-------------------------------

The downside of overloading the base print method is that the R CMD check returns
a warning. This means that *dfprint* will never make it to CRAN. If you want your
package to be on CRAN, importing *dfprint* is probably not an option.
On the other hand, if you are happy with people getting your package from GitHub,
you should be fine.

*dfprint* has no dependencies apart from what comes with base R (ie. *stats*
and *utils*). And while it is true that more dependencies is not always bad, it
is [worth thinking about](https://www.tidyverse.org/blog/2019/05/itdepends/). In
that regard, *dfprint* is a lean and leightweight choice.

Interactive use has much fewer constraints. I find myself using it all the time
when inspecting data frames, without necessarily including it as a dependency
in my projects.

```{r}
sessionInfo()
```
