---
title: "Printing data frames"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(width = 80)
```

*But data frames already have a print method!*

They do. But it falls short in a few ways which motivated people to reimagine
data frames altogether (AKA tibbles).

**1. Large data frames print poorly.** Tinkering with `[[` and `head()` is not
fun. Lists of data frames hanged R for me before when printed.

**2. Lack of metadata.** These are typically obtained by `str()`, `dim()` and
`object.size()`, at the cost of additional typing and scrolling, as the original
output gets displaced by subsequent calls.

**3. Too plain.** Less is not always more. Formatting would help to distinguish
factor from character, lists from atomic vectors etc.

**4. Valid inputs can produce an error.** For example:

```{r, error = TRUE}
li = list(as.symbol("a"), as.symbol("b"))
df = data.frame(a = 1:2)

df$li = li

df
```

The above is a data frame that cannot be printed because its list column `li`
includes symbol elements for which there is no formatting method. (EDIT: This
has been fixed in R 4.0.0.)

**5. And then there is this:** Data frames can have columns which themselves are
dataframes (or matrices or arrays). One place where you can see this is when you
parse JSON data structures, eg. using `jsonlinte::fromJSON()`. Can you tell from
the printout below which columns belong to the data frame column?

```{r}
df_parent = data.frame(x = 1:2, child.y = letters[1:2])
df_child  = data.frame(y = LETTERS[1:2], z = 2:3, stringsAsFactors = FALSE)

# Add child data frame as a column to the parent.
df_parent$child = df_child

# Check the number of columns
ncol(df_parent)

# Print out the result.
df_parent
```

There is no way of telling that the first `child.y` is a regular column, while
the rest belongs to the child data frame.

What to do
----------

I often find myself doing this:

```r
print.data.frame = function(df) print(tibble::as_tibble(df))
rm(print.data.frame) # when you no longer need it
```

But this will fail when POSIXlt columns are present, and will drop rownames.
Plus there is a risk that something unexpected may happen during the coercion to
tibble and ending up misinforming the user.

Overloaded!
-----------

So I wrote a package that overloads the `base::print.data.frame` method:

```{r}
library(dfprint)
```

This has its downsides, eg. it will never be on CRAN. But it makes it easier to
use - just load the package and your data frames will print nicely untill you
unload it or restart R.

How does it print?

```{r}
set.seed(1)
iris2 = iris[sample.int(nrow(iris)), ] #scramble the rows
iris2
```

Things to point out:

* The number of rows and columns is printed at the top, along with the amount
  of RAM allocated.
  
* Column headers include object type. For objects with a "class" attribute set
  (such as factors or datetimes), class is shown instead of type. Datetime
  columns (POSIXct and POSIXlt) also show the timezone.
  
* Factors display their underlying integer values to make their ordering a bit
  more obvious (see the `Species` column). This is useful for plotting.
  
* The footer is telling you how many rows are displayed. If you need more, use
  the `n` argument, eg, `print(iris2, n = 20)`. Use `Inf` to print all rows.
  
When a data frame is bit wider, the names and summaries of the overflowing
columns are displayed in the footer.

```{r}
set.seed(1)
# Produce more columns
iris3 = iris2[, sample.int(ncol(iris2), 8, replace = TRUE)]
iris3
```

Unlike with tibbles, rownames are preserved:

```{r}
mtcars[1:4, ]
```

And what about those failed examples from before? Here is the "unprintable"
data frame with some additional list columns added:

```{r}
df$li2 = list(mtcars, letters)
df$li3 = list(mean, ~x)
df
```

Prints out no problem. List columns include summaries of their elements instead
of the values themselves, which saves space and is often more informative.

Now for the data frame column - and let's also add a matrix column:

```{r}
df_parent$m = matrix(1:4, ncol = 2)
df_parent
```

The `$` and `[` make it obvious that these columns belong to 2-d structures. Note
that the number of columns stated above is correct, because `child` (not
`child.y`) and `m` are each a single 2-d column.

Options
-------

If your terminal does not fully support ANSI formatting, you can turn selected
features off.

```{r}
options(dfprint_strong = FALSE)

# Column headers are no longer bold:
mtcars
```

Use `dfprint_italic = FALSE` and `dfprint_subtle = FALSE` to disable individual
features, or `dfprint_enable = FALSE` to revert back to base print method.

Using dfprint in other packages
-------------------------------

The downside of overloading base print method is that the R CMD check returns
a warning. This means that *dfprint* will never make it to CRAN. If you want your
package to be on CRAN, importing *dfprint* is probably not an option.
On the other hand, if you are happy with people getting your package from GitHub,
you should be fine.

*dfprint* has no dependencies apart from what comes with base R (ie. *stats*
and *utils*). And while it is true that more dependencies is not always bad, it
is [worth thinking about](https://www.tidyverse.org/blog/2019/05/itdepends/). In
that regard, *dfprint* is a lean and leightweight choice.

I find myself using it all the time when inspecting data frames, without
necessarily including it as a dependency in my projects.
